Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-01 paper/current/main.tex
hierarchy result edits

The embedding space encodes the prestige hierarchy of U.S. universities that are coherent with real-world university rankings.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the correlation between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a bibliometrically-based university ranking is 0.87 (Spearman's $\rho$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding-based ranking and the Times ranking is robust regardless of the number of organizations used to define the axes (Fig. ~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities produces a ranking that is significantly correlated with the Times ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The correlation is also comparable to network-driven measures such as degree strength (sum of edge weights, Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see \SI). 
The strongest outliers that were ranked more highly in the Times ranking than in the embedding-based ranking tend to be large state universities such as Arizona State University and the University of Florida.
Those ranked higher in the embedding-based ranking tend to be relatively-small universities near major urban areas such as the University of San Francisco and the University of Maryland Baltimore County.
In sum, our result suggests that the embedding space is capable of capturing information about academic prestige, even when the presentation is learned using data without explicit information on the direction of mobility (as in other formal models~\autocite{clauset2015hierarchy}), or prestige.

The axes can further be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent, such as regional and liberal arts universities, seminaries, musical schools, community colleges (Fig.~\ref{fig:semaxis}d), government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).
Their estimatee prestige has a significant correlation with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}).

@@

We report that this curve is almost universal. For instance, China closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia, and other populous countries such as Brazil exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset), though their exact position and size vary.
Other nations exhibit different curves, such as Egypt which has an``S''-shaped curve, meaning that their largest organizations are not as well connected as other countries' largest organizations. 
Similar patterns can be found across many of the 30 of the countries with the most total researchers (Fig.~\ref{fig:concave30}), however, we can only speculate on what their variations mean (see \SI).
See Supporting Information for more discussion.

%
% Figure - Boomerang
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures prestige and size of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to the number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on the Times ranking.
		A convex-shape emerges, wherein larger universities tend to be more distant from the origin (large L2 norm); however, the more prestigious universities tend to have smaller L2 norms.
		The L2 norm is correlated weakly with the ratio of organizations' number of connections (non-zero flows) to the number of researchers ($R^{2} = 0.17$, \textbf{inset}).
		The red line is the line of the best fit, and 99\% confidence intervals are shown, though are too small to be visible.
		\textbf{b.} We find a similar convex-curve pattern across many countries such as the United States, China, Australia, Brazil, and others (\textit{inset}, and Fig.~\ref{fig:concave30}). 
		Some countries exhibit variants of this pattern, such as Egypt's case with missing the right side of the curve.
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries, with 99\% confidence intervals.
		Loess lines are also shown for organizations in Australia, Brazil, and Egypt (\textbf{inset}).
	}
	\label{fig:length}
\end{figure}

>>>>

The embedding space encodes the prestige hierarchy of U.S. universities that are coherent with real-world university rankings.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the correlation between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a bibliometrically-based university ranking is 0.87 (Spearman's $\rho$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding-based ranking and the Times ranking is robust regardless of the number of organizations used to define the axes (Fig. ~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities produces a ranking that is significantly correlated with the Times ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The correlation is also comparable to network-driven measures such as degree strength (sum of edge weights, Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see \SI). 
The strongest outliers that were ranked more highly in the Times ranking than in the embedding-based ranking tend to be large state universities such as Arizona State University and the University of Florida.
Those ranked higher in the embedding-based ranking tend to be relatively-small universities near major urban areas such as the University of San Francisco and the University of Maryland Baltimore County, possibly reflecting exchanges of scholars with nearby high-ranked institutions at these locations.
In sum, our result suggests that the embedding space is capable of capturing information about academic prestige, even when the representation is learned using data without explicit information on the direction of mobility (as in other formal models~\autocite{clauset2015hierarchy}), or prestige.

The axes can further be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent, such as regional and liberal arts universities, seminaries, musical schools, community colleges (Fig.~\ref{fig:semaxis}d), government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).
Their estimated prestige has a significant correlation with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}).

@@

We report that this curve is almost universal across many countries. 
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia, and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset), though their exact position and size vary.
Other nations exhibit different curves, such as Egypt which has an``S''-shaped curve, meaning that their largest organizations are not as well connected as other countries' largest organizations. 
Similar patterns can be found across many of the 30 of the countries with the most total researchers (Fig.~\ref{fig:concave30}), however, we can only speculate on what their variations mean (see \SI for more discussion).

%
% Figure - Boomerang
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures prestige and size of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to the number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on the Times ranking.
		A convex-shape emerges, wherein larger universities tend to be more distant from the origin (large L2 norm); however, the more prestigious universities tend to have smaller L2 norms.
		The L2 norm is correlated weakly with the ratio of organizations' number of connections (non-zero flows) to the number of researchers ($R^{2} = 0.17$, \textbf{inset}).
		The red line is the line of the best fit, and 99\% confidence intervals are shown, though are too small to be visible.
		\textbf{b.} We find a similar convex-curve pattern across many countries such as the United States, China, Australia, Brazil, and others (\textit{inset}, and Fig.~\ref{fig:concave30}). 
		Some countries exhibit variants of this pattern, such as Egypt, which is missing the right side of the curve.
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries, with 99\% confidence intervals.
		Loess lines are also shown for organizations in Australia, Brazil, and Egypt (\textbf{inset}).
	}
	\label{fig:length}
\end{figure}
